Synthesising Complex Termination Arguments

### 2014-06-18 papers/termination/paper.tex
revised intro + related work

\section{Introduction}

Proving program termination is typically done by finding a \emph{ranking function}
for the program states, i.e. a map from the program's state space to a well-ordered set.
As this definition of a ranking function is very general, research is often limited to some
convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.

Moreover, the majority of the existing termination provers diverge 
in one important aspect from the execution of a program on a computer:
bit-vectors and floats 
are treated as mathematical integers and reals, respectively.
This approximation gives rise to 
{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
Next, we will consider in more detail the assumptions made by such techniques and their implications.\\

\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
incomparable behaviours with respect to termination:
%The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
\begin{itemize}
\item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}
The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). 
\item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
 \begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
\end{itemize}

\noindent {\bf Floats vs. reals.} A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated next:
\begin{itemize}
\item This loop does not terminate for reals, but does for floats.
\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}

\item The following loop terminates for reals, but does not for floats.
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\end{itemize}   
\todo{explain the reasons for non-termination with figures.}\\

\noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
\emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
For this purpose, we rephrase the termination problem as a second-order satisfaction problem
(indeed, a program synthesis problem). 
As a direct consequence, our technique is semantic in nature and does not
dependent on the size of the analysed program or its number of variables. 
However, its efficiency does depend on the Kolmogorov
 complexity of the computed ranking functions. We show that the probability of
a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).

\begin{figure*}
\centering
 \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
 \hline
  & & \multicolumn{8}{c|}{Program} \\
  & & \multicolumn{2}{c|}{Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
  & & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13} & & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   & Non-linear non-lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
 \end{tabular}

@@

 The main contributions of our work can be summarised as follows:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We present a formulation that allows us to prove termination without using a fancy safety prover (because
we don't do a DWF proof).  In particular,
we only need a bounded model checker that performs a single unwinding of a loop.
\item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
multiple loops.
\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
\end{itemize}

@@

Other points to touch in the introduction:
\begin{itemize}
\item disjunctive well-foundedness: By phrasing the termination problem as a single rank function synthesis problem, we observe that
we can make the termination problem as hard as finding ranking functions, rather than as hard as
solving a difficult ``double reachability'' problem (cf. Terminator).
\item The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
\item Compared to the previous approaches that mostly aim to find a particular form of ranking functions, our technique does not commit to any particular one. We will find the one that has the lowest Kolmogorov complexity.
\item Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
reachable after leaving a previous loop.
\item Linearity vs. LKC.
\end{itemize}

This paper is organised as follows:

\section{Related Work}
\subsection{Termination analysis}
Given the multitude of works on termination, we have tried to summarise them in Figure~\ref{fig:handletable}.

The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.

A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
Consequently, the termination argument can be very simple. There are numerous techniques specialised in
proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.

Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.

There is very limited work on generating ranking functions for relations over machine integers.
Cook et al. present a method based on a reduction to Presburger
arithmetic, and a template-matching approach for predefined classes of
ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.

Leike and Heizmann present a new method for the constraint-based synthesis
of termination arguments for linear loop programs based on
linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.

Learning: \cite{DBLP:journals/corr/HeizmannHP14}

Polynomial programs: \cite{DBLP:conf/vmcai/BradleyMS05}

Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}.

\subsection{Disjunctive well-foundedness}
Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
%Terminator incrementally constructs the union of ranking relations.
%Given that the this union is not well-founded (it is only disjunctively well-founded), 
%Terminator must prove that the binary reachability relation of the program (the transitive closure of the transition relation restricted to reachable states) 
%is contained in the union of ranking relations.
%Thus, this approach requires an evolved binary reachability, and, implicitly, the assistance of a safety prover.
In particular, their technique incrementally constructs the union of ranking relations.
Given that the this union is not well-founded (it is only disjunctively well-founded),
they must show that, for every sequence, one of the ranking functions decreases
between the pre- and post-state. %In other words: it 
%must find all pairs of states reachable from the program's initial state 
%such that one is reachable from the other and it must show 
%that the value of one of the ranking functions decreases from s1 to s2. 
Consequently, most of the analysis time is spent in reachability analysis that is to check if the current
transition invariant reached a fixed point.

As we want to avoid the requirements of a binary reachability analysis, our termination argument is not a set
of ranking functions, but a single ranking function, which we prove is enough for proving the termination of any termination program. 
With a single ranking function we only need to show that the rank decreases from the
pre- to post-state after executing each single transition step.
The checking process is therefore easy. For the construction of the ranking function, we 
formulate the problem as a second order satisfaction formula and we solve it via program synthesis.

Because most state-of-the-art termination procedures use disjunctive well foundedness arguments, they require the use of
complex safety provers that are able to reason about unbounded unfoldings of a transition relation.  In contrast to this,
we are able to build monolithic termination arguments that only need ``local'' reasoning -- our ranking functions
can be proved sufficient by considering just a single unfolding of a loop.

@@

%-- the union T of ranking relations, however, is in general not well-founded
%(it is only disjunctively well-founded in the terminology of [23]). This is
%why it would not be sufficient to show the inclusion R \subseteq  T, and why we must in stead prove R_I^+ \subseteq T.

\subsection{Synthesis}

%% ---
%% constructs and checks the termination argument

%% T ERMINATOR’s most distinguishing aspect, with respect to pre-
%% vious methods and tools for proving program termination, is how it
%% shifts the balance between the two tasks of constructing and respec-
%% tively checking the termination argument. The classical method is
%% to construct an expression defining the rank of a state and then to
%% check that its value decreases in every transition from a reachable
%% state to a next one. The construction of the ranking function is the
%% hard part and forms a task that needs to be applied to the whole
%% program. The checking part is relatively easy. In our method, the
%% task of constructing ranking functions is the relatively easy part;
%% they are constructed on demand based on the examination of only
%% a few selected paths through the program.

%% Previously, it was not known whether binary reachability analysis could be made
%% practical. The challenge raised by our approach was to show that
%% this is indeed the case.

%% ----------
%% Efficient synthesis of ranking functions for machine-level bit-vectors, how-
%% ever, has remained an open problem.
%% ---------

@@

\subsection{Disjunctive Well-Foundedness}
Terminator uses disjunctive well foundedness arguments, which can be found
piece by piece, but require reasoning about the transitive closure of a transition
relation.  Because we build a monolithic well foundedness argument, we
do not need to reason about the full transitive closure, and can instead do our
reasoning over a single step of the transition relation.  This means that we
can get away with very simple model checking methods, rather than requiring
a full-blown safety prover.

@@

>>>>

\section{Introduction}\label{sec:intro}

Proving program termination is typically done by finding a \emph{ranking function}
for the program states, i.e. a map from the program's state space to a well-ordered set.
%% As this definition of a ranking function is very general, research is often limited to some
%% convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
%% In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
%% while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction. 
In doing so, the majority of the existing termination provers diverge 
in one important aspect from the execution of a program on a computer:
bit-vectors and floats are treated as mathematical integers and reals, respectively.
Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
most of them designed for integers and assuming a \emph{restrictive transition relation}, e.g. linear programs \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}. 
% \emph{linear ranking functions for linear programs over integers}.
These design decisions may give rise to 
{\em unsound} and {\em incomplete} analyses, as discussed next.
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

%\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} 
The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in 
incomparable behaviours with respect to termination. 
%The loop in Figure~\ref{} terminates for integers, but does not terminate for bit-vectors. 
The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). %This means that a termination analysis for integers would return an unsound result for this situation.
As a second example, the loop in Figure~\ref{} is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

%Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:

\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}

\begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}

%\noindent {\bf Floats vs. reals.} 
A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
While the former does not terminate for reals, but does for floats, the latter 
terminates for reals, but does not for floats.

\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\todo{explain the reasons for non-termination with figures.}\\

%% \noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
%% most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
%% \emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
%% function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.

If we survey the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve applying linear functions while corresponding global
measures involve nonlinear functions or lexicographic orders.

The downside of a Ramsey-based approach is the fact that a valid termination argument must hold for the \emph{transitive closure}
of the program's transitions, rather than only for individual transitions. 
As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 
requiring the support of powerful safety checkers.
Basically, these approaches opt for simpler termination arguments in the detriment of complex validity checking.

As Ramsey-based approaches are limited by the state of the art in safety checking, 
recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its very extreme: \emph{unrestricted} termination arguments. 
This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).
That is, we propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
checking the validity of a termination argument.

Given the level of generality that we aim for, we phrase the termination problem as a second order satisfaction problem (indeed, a program synthesis problem) (Section~\ref{}).
While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors influencing our method's performance and
find that our technique %is semantic in nature and 
does not dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
 complexity of the computed ranking functions. We show that the probability of
a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
Moreover, we find that our technique does actually gain from the generality of the termination argument. 
After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.

\begin{figure*}
\centering
 \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
 \hline
  & & \multicolumn{8}{c|}{Program} \\
  & & \multicolumn{2}{c|}{Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
  & & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
 \end{tabular}

@@

 The main contributions of our work can be summarised as follows:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
we only need a bounded model checker that performs a single unwinding of a loop.
%% \item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
%% multiple loops.
\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
\end{itemize}

@@

%semantic approach

%given that the search space size is constant, the probability of finding a solution 
%increases when considering all the possible solutions. There is no point in restricting the form of the solution..
%to read: constraint-based.., integers vs rationals.

%This paper is organised as follows:

\section{Related Work}
%\subsection{Termination analysis}
Automated program termination is a research topic that has received a fair amount of attention from the software verification community.
In order to compare our technique to the rest of the area, 
Figure~\ref{fig:handletable} summarises the related works with respect to the restrictions they impose on the transition relations as well as the form of the computed ranking functions.

Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}. 
Among them, 
Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
Leike and Heizmann present a new method for the constraint-based synthesis
of termination arguments for linear loop programs based on
linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
Learning: \cite{DBLP:journals/corr/HeizmannHP14}

%A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
%Consequently, the termination argument can be very simple. There are numerous techniques specialised in
%proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.

While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
there is very limited work for programs over machine integers.
Cook et al. present a method based on a reduction to Presburger
arithmetic, and a template-matching approach for predefined classes of
ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
Similarly, the only work we are aware of that can compute non-linear ranking functions  
for imperative loops with polynomial guards and polynomial assignments
is \cite{DBLP:conf/vmcai/BradleyMS05}.

Given the lack of research in handling \emph{not-linear programs}, as well as \emph{programs over bit-vectors and floats},  
our work focuses on covering these areas. 
One of the obvious conclusions that can be reached from observing Figure~\ref{fig:handletable}, 
is that most of the works tend to specialise on a certain aspect of termination proving that they can solve efficiently. 
Conversely to this view, we aim for \emph{generality}, as we do not restrict the form of the synthesised ranking functions, nor the form of the input programs.

As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's theorem compute a set of local termination conditions that decrease as execution follows through the loop
and require expensive reachability analyses
\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
In an attempt to reduce the complexity of checking the validity of the termination argument, 
Cook et al present an iterative termination proving procedure that searches for 
lexicographic termination arguments \cite{DBLP:conf/tacas/CookSZ13}, 
whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.

\todo{add smth about synthesis}

%The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
%For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
%with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.

@@

%\subsection{Synthesis}

@@

\subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
Terminator uses disjunctive well foundedness arguments, which can be found
piece by piece, but require reasoning about the transitive closure of a transition
relation.  Because we build a monolithic well foundedness argument, we
do not need to reason about the full transitive closure, and can instead do our
reasoning over a single step of the transition relation.  This means that we
can get away with very simple model checking methods, rather than requiring
a full-blown safety prover.

@@

Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
reachable after leaving a previous loop.
